A sustainable approach to tourist signage on heritage trails

### 2023-11-09 paper.tex
In seconda revisione

%=================================================================
% Full title of the paper (Capitalized)
\Title{A sustainable approach to tourist signage on heritage trails\footnote{This research has been supported by the Ministry of University and Research under the "Underlandscape" project (2020428LS8)}}
% MDPI internal command: Title for citation in the left column
\TitleCitation{A sustainable approach to tourist signage on heritage trails}
% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0009-0003-1136-570X} % Add \orcidA{} behind the author's name
\newcommand{\orcidauthorB}{0000-0002-9734-2044} % Add \orcidA{} behind the author's name
\newcommand{\orcidauthorC}{0000-0002-5542-7891} % Add \orcidB{} behind the author's name

@@

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Guiding the visitor to appreciate the sights during an excursion is crucial to building a memorable experience. In the literature, we find many articles exploring the use of sophisticated technologies towards such a goal. The case of tourism proposals for inland areas exhibits a noteworthy aspect, since the presence of human artifacts and the signage may damage the experience, and introduce pollutants as well. This paper analyzes the signage for a natural trail putting into practice a holistic methodology that takes into account all aspects of the problem, starting from the touristic vocation of the area and the social fabric. The analysis indicates slow, community-involved tourism as the preferable target, with extreme requirements of sustainability in the social, economic, and environmental domains for the signage solution. Appropriate governance guidelines are given to implement the proposal. Footprint evaluations address the QR-code technology, which is used to design a real-scale solution finally deployed on the field.}
%\abstract{A single paragraph of about 200 words maximum. For research articles, abstracts should give a pertinent overview of the work. We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations. The abstract should be an objective representation of the article, it must not contain results that are not presented and substantiated in the main text and should not exaggerate the main conclusions.}

@@

Although an official decision is still pending, there is a widely agreed proposal for the naming of the time that we are living as the Anthropocene.  Behind this proposal is the recognition that humans have the potential to significantly alter on a global scale the environment in which they live.

@@

Leisure and cultural activities, like tourism, bring profit to niche businesses. This creates a reinforcing effect that follows the negative dynamics seen above: a activity that is neutral in itself (e.g., skying) can have a significant impact due to induced effects (lifts) that strengthen the success of the activity (easier to practice) bringing more investment and impact. Unlike the beneficial effects, which are usually well localized, the negative impact has a very broad spectrum that includes environmental, social, and economic sectors, as discussed by Patthey in \cite{pat08a} relating to outdoor winter sports.

@@

This research stems from such a tough contradiction in trying to answer the question: is there a way to unlock the potential of cultural heritage while at the same time limiting the damage to a social, economic, and environmental equilibrium?

@@

This paper illustrates a concrete exercise in designing for sustainability. We introduce an instance of cultural sustainability, an inland region with a legacy to defend, and design a concrete tool to address the issue. To reach fully sustainable result, we need to practice from the early design stages an holistic approach to the environmental, social and financial facets of the problem.

@@

The next section illustrates signage as a topic of interest. The discussion targets low cost, low impact technologies that are appropriate for outdoor signage of natural or cultural heritage, and finally indicates QR code tags as a consistent technology. The section features a survey of relevant literature.

Section \ref{sec:methods} provides the multidisciplinary framework. Section \ref{sec:legacy} illustrates the legacy emphasizing historical facts and locations. Section \ref{sec:tourism} describes the social and touristic context of the target location and provides the guidelines for integrating the initiative in the socio-political fabric. Such guidelines are used to motivate and define an agenda, in section \ref{sec:agenda}, and the section is concluded by a reasoned description of the area, highlighting the points of interest and their reachability \ref{sec:soa}.

Section \ref{sec:implementation} describes the concrete implementation that follows the defined requirements. The description of the process to produce the printed tags starting from editable content on a geographic database allows to reproduce and reuse the solution in other contexts.

Section \ref{sec:results} merges the various aspects explored in the previous sections to provide a methodological view of the process, focusing on the interplay between the fourth pillars of sustainability.

@@

The functions of signage are two:
\begin {itemize}
\item an effective signage must guide the tourist across the resort. A map represents a starting point for this purpose, but not all visitors feel comfortable reading a map. The ideal signage should provide visible reference points and suggest actions, like "turn right after crossing the stream". A good example of such signage is the one used by the Italian Alpine Club (CAI), consisting of colored stripes marked on tree bark (see figure \ref{fig:CAI}). The marks are placed in such a way that standing near one of them, the hiker can spot the successive on the trail;
 \item a sign should explain the reasons for interest in a site. This generic characterization includes all sorts of semantic characterization of the site, and may have educational or motivational intents. This includes describing historical facts related with the site, as well as advertising nearby points of interest (see figure \ref{fig:pole-direction}). In this sense, the CAI signal above is not sufficient. A simple board with a site name fits only when the site is sufficiently popular. Otherwise, a more structured message should be used to pursue an educational intent.
\end{itemize}

@@

In our study, we consider a range of solutions that support sustainable tourism, specifically the dynamic provision of information during a visit along a natural trail not covered by Internet connectivity. We consider the presence of two stakeholders for the signage itself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts useful and enjoyable information from signing devices.

@@

One is that smartphones although widely available, are not equally familiar to everybody. This is related to {\em usability}, with a term borrowed from P. Wan in \cite{wan22a}. The second is that several functions depend on the provision of enabling services: for instance, their networking capability is useless if the device cannot reach the Internet. So the {\em applicability limits} of a solution need to be defined.

Starting from the two aspects above we envision the guidelines for a successful smartphone-based strategy, keeping in mind its sustainability.

@@

A popular research direction covers the social aspects related to the use of the smartphones. W. Tan \cite{tan17a} covers all aspects of a touristic experience related to the smartphone, from the definition of travel destination to assistance during the visit. Much attention is dedicated to the network of connections that is established thanks to the smartphone, which, again, is considered on the Internet. Such an assumption is in contrast with the title, which indicates a nature-based destination, where notably the Internet is not always reachable.

@@

More recently, the smartphone has been considered not strictly related to communication. In 2021 A. Slavec et al. investigated the use of cameras \cite{sla21a} while on travel in locations with a relevant cultural heritage to sustain its preservation and engage the tourists using location-based games, similar to Pokemon Go or geo-caching.

In 2023 V. Rodrigues et al. published a systematic review of papers considering the interrelationships between tourism and portable digital devices \cite{rod23a}. Although the title evokes a one-way contribution, i.e., the impact of digitalization is assumed to be positive on the quality and the sustainability of touristic offers, in the conclusions the authors reveal the awareness for the need to address {\em "the preservation of tourism attractions/sites"} and call for a {\em "a holistic approach ... to support a concept that still lacks conceptual and empirical clarification"}.

@@

Regarding the specific technology we refer to, QR codes have been invented in 1994 for industrial applications. The current version, described in \cite{isoqr}, is dated February 2015. In 2013 an early paper by Martinez et al. \cite{mar13a} suggests its utilization to implement a virtual tour in the {\em Las Quilamas Natural Park} in Spain. At that time the availability of QR-code readers was not as diffused as today, and therefore the work may be considered among the seminal ones. The authors proposed to use the QR code as a link to a panel in PDF format viewed on the user's portable device. A similar work in a urban environment is presented in \cite{fin13a} and \cite{tat15a}, the latter illustrating the QR system serving the monuments in the Serbian town of Niš.

Since then the use of the QR code technology for heritage promotion has been suggested in other papers. In \cite{sol18a} the authors describe its application in a museum: after defining the points of interest, tags have been placed and linked to related documents. The authors conducted a poll about user perception on two segments of users: Italians and Polish. The QR tags were not unanimously considered useful: in particular half of the Italian users considered them as not useful. It is interesting to note that the motivation of the other segment of users, less skeptical about usefulness, shared a solid educational motivation to access QR-linked content (70\% in the Polish group against 50\% in the Italian one). This fact sheds light on the necessity to provide content the user needs: curiosity is not a sufficient motivation.

@@

There are examples of urban outdoor applications of QR tags, like

All the found publications share a common attitude, which can be summarized in the motto "analog portal to digital world", as in the title of \cite{bai12a}. None of them evaluates the possibility of storing text in the tag itself, thus eliminating the requirement of an Internet connection.

The present paper promotes the use of QR codes with sustainability and as the primary or only way of providing information to the visitor. In the literature, QR codes are considered ancillary to other forms of communication, like physical installations or the Web. Therefore, in our case the informational content of the tag, which is commonly ignored, becomes relevant. It is a new and original way of using QR codes for the delivery of information.

We emphasize that this article is not meant to quantify user satisfaction or the economic revenue associated with the specific solution. Such a target is outside the scope of our research, and indeed the figures that would measure the success of a strategy deserve a thorough investigation. We aim at isolating an issue, proposing a sustainable strategy for its solution, and implementing a proof of concept for it.

@@

The wireless capabilities of a smartphone are the focus of a straightforward approach. Given the premise that the Internet is assumed to be unreachable, the {\em host} stakeholder might provide a local network of low-power radios covering the region of interest. Small servers connected to the network would host specific Web services. The approach requires a modest investment and a marginal environmental impact: for instance, a small device based on the ESP8266 single-chip computer (SCC) has a coverage of tens of meters, and a volume in the order of the $cm^3$. It can provide a WiFi Access Point as well as Web content. The ESP8266 has a capacity of 32KB, which is sufficient for explanatory text and a low-resolution image. Other SCCs, like the ESP32, are more powerful but exhibit a higher power consumption.

Such a solution is severely limited by the dependency from a power supply. A radio transmitter is a rather power-consuming device, in the range of Watts. Even if intermittent, the operation of a battery-operated networking device cannot be guaranteed for long periods and the host organization should consider power harvesting, which negatively contributes to the economic cost and environmental footprint of the device.

@@

The NFC technology is available without installing ad-hoc applications. Its operation consists in approaching the smartphone to the tag, and the user is advertised that a content is available. The smartphone can read it aloud to compensate for the user's inability or translate to deal with linguistic issues. Such capabilities do not need an Internet connection but currently depend on ad-hoc applications.

Another solution in the family of passive devices is the QR tag (QR stands for Quick-Response). Such a technology does not require radio communication but uses the smartphone camera to acquire text encoded in a graphical code. The capacity of a QR tag depends on the number of dots in the image, which in turn depends on the size of the code and the smartphone camera characteristics. We may consider that the capacity of a QR tag roughly equals that of an NFC tag. A QR tag is larger than an NFC tag of comparable capacity, and manufacturing requires a printer.

@@

\begin{itemize}
	\item {\em Equitable use} is closely related to the smartphone technology, which is itself considered a vehicle for equitability,
	\item {\em Flexibility in use} is enabled by the device capabilities, which allow listening instead of reading, translating the information in a different language, or storing it for later use,
	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single finger touch on the smartphone,
	\item{\em Perceptible information} is a critical requirement, which contrasts to keep low environmental intrusion. This point will be further discussed in the section devoted to the implementation,
	\item{\em Tolerance for error}: there is no chance of accidentally causing a hazardous situation using the QR tag. The unintended release of the passive device into the environment determines a minor pollution,
	\item{\em Low physical effort} holds, although the user needs to carry a smartphone,
	\item{\em Size and Space for approach and use} need to be carefully considered. In the case of the NFC tag the smartphone needs to be nearly in touch with the passive device, while the QR code must be in the line of sight and frameable without effort.
\end{itemize}

Compared with other smartphone-based technologies, there is a trade-off concerning capacity. However, a signage applications a capacity of 2-300 words is sufficient to provide directions for the visit and convey text-based educational contents. If capacity limits are not an issue, a solution based on NFC or QR tags exhibits several advantages:

\begin{itemize} 
\item does not require power supply,
\item has a limited impact on the landscape,
\item does not entice theft,
\item has a negligible cost,
\item is durable,
\item produces a limited quantity of waste when disposed of,
\item content can be stored for later usage; for instance, to visit a URL once the user reaches a zone covered by the Internet,
\item is inclusive regarding user inabilities and nationality,
\item requires basic technical skills from the user.
\end{itemize}

@@

Its use is especially appropriate when the cultural heritage extends across a scarcely accessible land, preserving and accessing sites located on private property or in hazardous areas can be challenging. In these cases, QR tags could improve the visitors’ experience by providing, besides textual data, the Web URLs for videos, aerial shots, and photos, even if they don’t have access to a specific spot. Multimedia resources allow visitors to immerse themselves in the area's history and gain an experiential understanding, even if they don't have access to a specific spot. The resulting widespread knowledge can be a beneficial tourist-economic engine that encourages investments in cultural heritage policies, thus achieving the goal of protecting and conserving abandoned sites.

@@

Ceramic fragments dated between the 3rd century BC and the 1st-2nd century AD prove that Ligurian populations occupied the region scattered and in small nuclei. The dedication of the Latin colony of \textit{Lucca} in 180 BC marked a decisive turning point in the Romanisation of the area and, shortly after, the definitive subjugation of the Ligurian populations, accompanied by a rapid acculturation \cite{gia96, cia05}.

@@

The Tourism Density Index is the rate between the number of tourists and the surface area in the higher touristic season. It is 106.41 tourists per $km^2$ in 2022. The arrivals exhibit a drop from 2019 (from  4339 in 2017 to 2363 in 2021), and touristic presences have suffered a slight decrease during the same period, as shown in figure \ref{fig:tourismflows}, while a further increase has been registered in 2022 (5.876 arrivals).

@@

As shown in \ref{tab:tourism}, the number of nights per tourist amounts to an average of 2,98 nights per tourist during a year and it is also significant since it shows, on one hand, \emph{Bagni di Lucca} context appeal for tourists choosing to stay a number of nights more than a weekend and less than a week. On the other hand, this data means that people like staying in this tourist comprehensory and they may have found attractions, services and activities they need. Consequently, a well-organized touristic system attracts potentially touristic presences widespread on the municipality territory.

@@

For a benchmarking point of view, the provincial touristic context of \emph{Lucca} is mentioned in this study with the purpose of highlighting the percentage weight of \emph{Bagni di Lucca} tourism flows within the whole tourism dynamics of the Province of \emph{Lucca}. 
As concerns tourist indexes shown in Table \ref{tab:tourism}, the touristic density counts 1866.96 people per square kilometer (1773 $km^2$), with an average stay of 3.36 nights per tourist, showing a medium-length permanence in the provincial territory; this data is in line with \emph{Bagni di Lucca} average stay (2.98 nights). 
The provincial area of \emph{Lucca} shows an accommodation density of 19.30 on 10.14 of \emph{Bagni di Lucca}, while the coexistence index of \emph{Lucca} (with 69.87 foreign tourists on 100 Italian tourists) is lower than \emph{Bagni di Lucca}’s coexistence index (with 103.74 foreign tourists in 100 Italian tourists). 
Definitely, it can be claimed that \emph{Bagni di Lucca} is well positioned in the whole touristic context of \emph{Lucca} since it represents one of the less populated municipalities in the territory with interesting tourist indexes. In fact, \emph{Bagni di Lucca} boasts its historical thermal tourism tradition together with a well-equipped environment for sport and adventure tourism, on a surface of 164.70 $km^2$, that is one of the biggest municipality areas in the province of \emph{Lucca}.

@@

\emph{Casoli} village boasts a heterogeneous range of tourist attractions, including both soft and hard outdoor experiential activities, such as: canyoning, rafting, and trekking experiences in the Canyon Park. Additionally, it features Romanesque churches and huts attracting visitors looking for various experiences such as thermal, environmental, outdoor, sport and cultural activities. Throughout the year, local stakeholders offer a variety of services, often incorporating sustainable tourism practices. After a detailed analysis of the tourist and social context of \emph{Bagni di Lucca}, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of touristic experiences \cite{loz09} dating back to the ‘90s, when some associations began organizing mainly trekking excursions and rafting experience .

As part of \emph{Bagni di Lucca}'s ecosystem, with its focus on tourism, \emph{Casoli} is a destination for slow travel. It is important to ensure that any development actions taken are sustainable, in order to preserve its unique social fabric. This means paying attention to conservation of the environment and cultural heritage, as well as preserving intergenerational continuity and economic equity \cite{kuh10a}.

@@

Based on the tourism trend of \emph{Casoli} and its overall tourism, and in alignment with the research goals focused on sustainable tourism, it is crucial to consider the role of local policy-makers. In fact, they play a vital role in planning sustainable measures and monitoring their impact.

In these terms, as described by [50] Dianne Dredge, and in light of the above analysis, private and public entities can contribute to sustainable tourism by developing the following actions:

@@

In order to create a successful strategic plan, it is crucial to have a thorough understanding of the destination's morphology, environment, history, culture, society, and economy, with sustainability being the key value.
Policymakers should recognize such needs and act in this way, both politically and socially as described by Beatrix et al. in \cite{bea10}. As concerns the empiric case of \emph{Casoli} village, within the wider context of \emph{Bagni di Lucca}, tourism is a key driver for re-population and re-qualification strategies, mainly where historical and archaeological sight can be promoted with natural attractions.

@@

\subsubsection{How top implement the guidelines? \label{sec:agenda}}

We include three stakeholder communities in the evaluation of the response to the initiative. We do not speak about {\em success} since the initiative is deliberately soft, an outcome expected in years, and the definition of the evaluation criteria raises new questions. We are content with determining if the stakeholders viewed the initiative as intrusive, helpful, or simply neutral.

@@

\begin{itemize}
	\item {\em the residents} the local community that currently inhabits the site
	\item {\em the entrepreneurs} who currently have a business on the site
	\item {\em the administrators} that are in charge of managing the site resources and that, at due time, will respond to the two stakeholders above, and finally
	\item {\em the users}, those that come to \emph{Casoli} to visit the Cave of La Piella and that we find the QR-codes on their way
\end{itemize}

The QR tag does not record reading operations, preventing an evaluation of the user category. A URL was added to the text message in the QR tag to monitor the number of hits.: however, this feedback largely underestimates the number of times the tag is read.

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{A sustainable approach to tourist signage on heritage trails
%\footnote{This research has been supported by the Ministry of University and Research under the "Underlandscape" project (2020428LS8)}
}
% MDPI internal command: Title for citation in the left column
\TitleCitation{A sustainable approach to tourist signage on heritage trails}
% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0009-0003-1136-570X} % Add \orcidA{} behind the author's name
\newcommand{\orcidauthorB}{0000-0002-9734-2044} % Add \orcidA{} behind the author's name
\newcommand{\orcidauthorC}{0000-0002-5542-7891} % Add \orcidB{} behind the author's name

@@

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Understanding the cultural aspects of an area rich in heritage is crucial to building a lasting educational experience from an excursion. Many articles in literature explore the use of sophisticated technologies to achieve such a goal. Tourism proposals for inland areas are significant because the presence of human artifacts and signage can harm the experience and create pollutants. Through a holistic methodology and analysis, this paper examines the signage for an area rich in cultural and natural assets: the study encompasses history, touristic vocation, and the environmental context. According to the analysis, slow, community-involved tourism is the preferred destination, and the signage solution must meet strict sustainability requirements in the social, economic, and environmental realms. After applying the appropriate governance guidelines provided, the QR code technology is selected for a thoroughly documented experimental deployment.}
%\abstract{A single paragraph of about 200 words maximum. For research articles, abstracts should give a pertinent overview of the work. We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations. The abstract should be an objective representation of the article, it must not contain results that are not presented and substantiated in the main text and should not exaggerate the main conclusions.}

@@

Although an official decision is still pending, there is a widely agreed proposal for the naming of the time that we are living as the Anthropocene.  Behind this proposal is the recognition that humans have the potential to significantly alter, on a global scale, the environment in which they live.

@@

Leisure and cultural activities, like tourism, bring profit to niche businesses. This creates a reinforcing effect that follows the negative dynamics seen above: an activity that is neutral in itself (e.g., skying) can have a significant impact due to induced effects (lifts) that strengthen the success of the activity (easier to practice) bringing more investment and impact. Unlike the beneficial effects, which are usually well localized, the negative impact has a very broad spectrum that includes environmental, social, and economic sectors, as discussed by Patthey in \cite{pat08a} relating to outdoor winter sports.

@@

This research stems from such a tough contradiction in trying to answer the question: which is the way to unlock the potential of cultural heritage while at the same time limiting the damage to a social, economic, and environmental equilibrium?

@@

This paper illustrates a concrete exercise in designing for sustainability. We introduce an instance of cultural sustainability, an inland region with a legacy to defend, and design a concrete tool to address the issue. To reach fully sustainable results, we need to practice from the early design stages a holistic approach to the environmental, social, and financial facets of the problem.

@@

The next section illustrates signage as a topic of interest. The discussion targets low-cost, low-impact technologies that are appropriate for outdoor signage of natural or cultural heritage, and finally indicates QR code tags as a consistent technology. The section features a survey of relevant literature.

Section \ref{sec:methods} provides the multidisciplinary framework. Section \ref{sec:legacy} illustrates the legacy emphasizing historical facts and locations. Section \ref{sec:tourism} describes the social and touristic context of the target location and provides the guidelines for integrating the initiative into the socio-political fabric. Such guidelines are used to motivate and define an agenda, in section \ref{sec:agenda}, and the section is concluded by a reasoned description of the area, highlighting the points of interest and their reachability \ref{sec:soa}. Section \ref{sec:implementation} describes the concrete implementation that follows the defined requirements. The description of the process to produce the printed tags starting from editable content on a geographic database allows reproduction and reuse of the solution in other contexts.

Section \ref{sec:results} merges the various aspects explored in the previous sections to provide a methodological view of the process, focusing on the interplay between the four pillars of sustainability.

@@

The functions of signage are two:
\begin {itemize}
\item an effective signage must guide the tourist across the resort. A map represents a starting point for this purpose, but not all visitors feel comfortable reading a map. The ideal signage should provide visible reference points and suggest actions, like "turn right after crossing the stream". A good example of such signage is the one used by the Italian Alpine Club (CAI), consisting of colored stripes marked on tree bark (see figure \ref{fig:CAI}). The marks are placed in such a way that standing near one of them, the hiker can spot the successive on the trail;
 \item a sign should explain the reasons for interest in a site. This generic characterization includes all sorts of semantic characterization of the site and may have educational or motivational intents. This includes describing historical facts related to the site, as well as advertising nearby points of interest (see figure \ref{fig:pole-direction}). In this sense, the CAI signal above is not sufficient. A simple board with a site name fits only when the site is sufficiently popular. Otherwise, a more structured message should be used to pursue an educational intent.
\end{itemize}

@@

In our study, we consider a range of solutions that support sustainable tourism, specifically the dynamic provision of information during a visit along a natural trail not covered by Internet connectivity. We consider the presence of two stakeholders for the signage itself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts information from signing devices.

@@

One is that smartphones, although widely available, are not equally familiar to everybody. This is related to {\em usability}, with a term borrowed from P. Wan in \cite{wan22a}. The second is that several functions depend on the provision of enabling services: for instance, their networking capability is useless if the device cannot reach the Internet. So the {\em applicability limits} of a smartphone-based solution need to be defined.

Starting from the two aspects above we envision the guidelines for a successful strategy, keeping in mind its sustainability.

@@

A popular research direction covers the social aspects related to the use of smartphones. W. Tan \cite{tan17a} covers all aspects of a touristic experience related to the smartphone, from the definition of travel destination to assistance during the visit. Much attention is dedicated to the network of connections that is established thanks to the smartphone, which, again, is on the Internet. Such an assumption is in contrast with the title, which indicates a outdoor-based destination, where notably the Internet is not always reachable.

@@

More recently, the smartphone has been considered not strictly related to communication. In 2021 A. Slavec et al. investigated the use of cameras \cite{sla21a} while on travel in locations with a relevant cultural heritage to sustain preservation and engage the tourists using location-based games, similar to Pokemon Go or geo-caching.

In 2023 V. Rodrigues et al. published a systematic review of papers considering the interrelationships between tourism and portable digital devices \cite{rod23a}. Although the title evokes a one-way contribution, therefore assuming digitalization improves quality and sustainability of touristic offers, in the conclusions the authors reveal the awareness for the need to address {\em "the preservation of tourism attractions/sites"} and call for a {\em "a holistic approach ... to support a concept that still lacks conceptual and empirical clarification"}.

@@

Regarding the QR code, the specific technology are going to use, it have been invented in 1994 for industrial applications. The current version, described in \cite{isoqr}, is dated February 2015. In 2013 an early paper by Martinez et al. \cite{mar13a} suggests its utilization to implement a virtual tour in the {\em Las Quilamas Natural Park} in Spain. At that time the availability of QR-code readers was not as diffused as today, and therefore the work may be considered among the seminal ones. The authors proposed to use the QR code as a link to a panel in PDF format viewed on the user's portable device. A similar work in an urban environment is presented in \cite{fin13a} and \cite{tat15a}, the latter illustrating the QR system serving the monuments in the Serbian town of Niš.

Since then, the use of the QR code technology for heritage promotion has been suggested in other papers. In \cite{sol18a} the authors describe its application in a museum: after defining the points of interest, tags have been placed and linked to related documents. The authors conducted a poll about user perception on two segments of users: Italians and Polish. The QR tags were not unanimously considered useful: in particular half of the Italian users considered them as not useful. It is interesting to note that the motivation of the other segment of users, less skeptical about usefulness, shared a solid educational motivation to access QR-linked content (70\% in the Polish group against 50\% in the Italian one). This fact sheds light on the necessity to provide content the user needs: curiosity is not a sufficient motivation.

@@

All the referenced publications share a common attitude, which can be summarized in the motto "analog portal to the digital world", as in the title of \cite{bai12a}. None of them evaluates the possibility of storing text in the tag itself, thus eliminating the requirement of an Internet connection.

The present paper promotes the use of QR codes as the primary or only way of providing information to the visitor. Unlike in the literature, where QR tags are considered ancillary to other forms of communication, like physical installations or linked Web contents. Therefore, in our case the informational content of the tag, which is commonly ignored, becomes relevant. It is a new and original way of using a QR code for the delivery of information.

%We emphasize that this article is not meant to quantify user satisfaction or the economic revenue associated with the specific solution. Such a target is outside the scope of our research, and indeed the figures that would measure the success of a strategy deserve a thorough investigation. We aim at isolating an issue, proposing a sustainable strategy for its solution, and implementing a proof of concept for it.

@@

The wireless capabilities of a smartphone are the focus of a straightforward approach which however shows critical issues. Given the premise that the Internet is assumed to be unreachable, the {\em host} stakeholder might provide a local network of low-power radios covering the region of interest. Small servers connected to the network would host specific Web services. The approach requires a modest investment and a marginal environmental impact: for instance, a small device based on the ESP8266 single-chip computer (SCC) has a coverage of tens of meters, and a volume in the order of the $cm^3$. It can provide a WiFi Access Point as well as Web content. The ESP8266 has a capacity of 32KB, which is sufficient for explanatory text and a low-resolution image. Other SCCs, like the ESP32, are more powerful but exhibit a higher power consumption.

Such a solution is severely limited by the dependency on a power supply. A radio transmitter is a rather power-consuming device, in the range of Watts. Even if intermittent, the operation of a battery-operated networking device cannot be guaranteed for long periods and the host organization should consider power harvesting, which negatively contributes to the economic cost and environmental footprint of the device.

@@

Reading a tag does not require an ad-hoc application, and is performed approaching the smartphone to the tag: the user is advertised that content is available. The smartphone can read it aloud to compensate for the user's inability or translate to deal with linguistic issues. Such capabilities do not need an Internet connection but currently depend on ad-hoc applications.

Another solution in the family of passive devices is the QR code (QR stands for Quick-Response). Such a technology does not require radio communication but uses the smartphone camera to acquire text encoded in a graphical code. The capacity of a QR tag depends on the number of dots in the image, which in turn depends on the size of the code and the smartphone camera characteristics. We may consider that the capacity of a QR tag roughly equals that of an NFC tag. A QR tag is larger than an NFC tag of comparable capacity, and manufacturing requires a printer.

@@

\begin{itemize}
	\item {\em Equitable use} is closely related to the smartphone technology, which is itself considered a vehicle for equitability,
	\item {\em Flexibility in use} is enabled by the device capabilities, which allow listening instead of reading, translating the information in a different language, or storing it for later use,
	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single finger touch on the smartphone,
	\item{\em Perceptible information} is a critical requirement, which contrasts with the need to keep the low environmental intrusion. This point will be further discussed in the section devoted to the implementation,
	\item{\em Tolerance for error}: there is no chance of accidentally causing a hazardous situation using the QR tag. The unintended release of the passive device into the environment determines a minor pollution,
	\item{\em Low physical effort} holds, although the user needs to carry a smartphone,
	\item{\em Size and Space for approach and use} need to be carefully considered. In the case of the NFC tag the smartphone needs to be nearly in touch with the passive device, while the QR code must be in the line of sight and frameable without effort.
\end{itemize}

Compared with other smartphone-based technologies, there is a trade-off concerning capacity. However, a signage application with a capacity of 2-300 words is sufficient to provide directions for the visit and convey text-based educational content. If capacity limits are not an issue, a solution based on NFC or QR tags exhibits several advantages:

\begin{itemize} 
\item does not require a power supply,
\item has a limited impact on the landscape,
\item does not entice theft,
\item has a negligible cost,
\item is durable,
\item produces a limited quantity of waste when disposed of,
\item content can be stored for later usage; for instance, to visit a URL once the user reaches a zone covered by the Internet,
\item is inclusive regarding user inabilities and nationality,
\item requires only basic technical skills from the user.
\end{itemize}

@@

Its use is especially appropriate when the cultural heritage extends across scarcely accessible land, considering that preserving and accessing sites located on private property or in hazardous areas can be challenging. In such cases, QR tags could improve the visitors’ experience by providing, when Internet coverage is available, the Web URLs for videos, aerial shots, and photos, even if the specific spot is not accessible. Such resources allow visitors to immerse themselves in the area's history and gain an experiential understanding.

@@

Ceramic fragments dated between the 3rd century BC and the 1st-2nd century AD prove that Ligurian populations occupied the region scattered and in small nuclei. The dedication of the Latin colony of \textit{Lucca} in 180 BC marked a decisive turning point in the Romanisation of the area and, shortly after, the definitive subjugation of the Ligurian populations, accompanied by rapid acculturation \cite{gia96, cia05}.

@@

The Tourism Density Index is the rate between the number of tourists and the surface area in the higher tourist season. It is 106.41 tourists per $km^2$ in 2022. The arrivals exhibit a drop from 2019 (from  4339 in 2017 to 2363 in 2021), and touristic presences have suffered a slight decrease during the same period, as shown in figure \ref{fig:tourismflows}, while a further increase has been registered in 2022 (5.876 arrivals).

@@

As shown in \ref{tab:tourism}, the number of nights per tourist amounts to an average of 2,98 nights per tourist during a year and it is also significant since it shows, on one hand, \emph{Bagni di Lucca} context appeal for tourists choosing to stay more than a weekend and less than a week. On the other hand, this data means how long people likes staying in this tourist region and that they may have found attractions, services, and activities they need. Consequently, a well-organized touristic system attracts potentially touristic presences widespread on the municipality territory.

@@

For a benchmarking point of view, the provincial touristic context of \emph{Lucca} is mentioned in this study to highlight the percentage weight of \emph{Bagni di Lucca} tourism flows within the whole tourism dynamics of the Province of \emph{Lucca}. 
As concerns tourist indexes shown in Table \ref{tab:tourism}, the touristic density counts 1866.96 people per square kilometer (1773 $km^2$), with an average stay of 3.36 nights per tourist, showing a medium-length permanence in the provincial territory; this data is in line with \emph{Bagni di Lucca} average stay (2.98 nights). 
The provincial area of \emph{Lucca} shows an accommodation density of 19.30 on 10.14 of \emph{Bagni di Lucca}, while the coexistence index of \emph{Lucca} (with 69.87 foreign tourists on 100 Italian tourists) is lower than \emph{Bagni di Lucca}’s coexistence index (with 103.74 foreign tourists in 100 Italian tourists). 
It can be claimed that \emph{Bagni di Lucca} is well positioned in the whole touristic context of \emph{Lucca} since it represents one of the less populated municipalities in the territory with interesting tourist indexes. In fact, \emph{Bagni di Lucca} boasts its historical thermal tourism tradition together with a well-equipped environment for sport and adventure tourism, on a surface of 164.70 $km^2$, that is one of the biggest municipality areas in the province of \emph{Lucca}.

@@

\emph{Casoli} village boasts a heterogeneous range of tourist attractions, including both soft and hard outdoor experiential activities, such as canyoning, rafting, and trekking experiences in the Canyon Park. Additionally, it features Romanesque churches and huts attracting visitors looking for various experiences such as thermal, environmental, outdoor, sports, and cultural activities. Throughout the year, local stakeholders offer a variety of services, often incorporating sustainable tourism practices. After a detailed analysis of the tourist and social context of \emph{Bagni di Lucca}, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of touristic experiences \cite{loz09} dating back to the '90s when some associations began organizing mainly trekking excursions and rafting experience.

As part of \emph{Bagni di Lucca}'s ecosystem, with its focus on tourism, \emph{Casoli} is a destination for slow travel. It is important to ensure that any development actions taken are sustainable, to preserve its unique social fabric. This means paying attention to the conservation of the environment and cultural heritage, as well as preserving intergenerational continuity and economic equity \cite{kuh10a}.

@@

Based on the tourism trend of \emph{Casoli}, and aligned with the research goals focused on sustainability, it is crucial to consider the role of local policy-makers. They play a vital role in planning and monitoring sustainable measures and their impact.

In these terms, as described by \cite{dre22} Dianne Dredge, and in light of the above analysis, private and public entities can contribute to sustainable tourism by developing the following actions:

@@

To create a successful strategic plan, it is crucial to have a thorough understanding of the destination's morphology, environment, history, culture, society, and economy, with sustainability being the key value.
Policymakers should recognize such needs and act in this way, both politically and socially as described by Beatrix et al. in \cite{bea10}. As concerns the empiric case of \emph{Casoli} village, within the wider context of \emph{Bagni di Lucca}, tourism is a key driver for re-population and re-qualification strategies, mainly where historical and archaeological sight can be promoted with natural attractions.

@@

\subsubsection{How to implement the guidelines? \label{sec:agenda}}

According with the above guidelines for tourism valorization, we explain how the research has been conducted dealing with local stakeholders. Although information is still incomplete, we were able to collect sufficient opinions to confirm that \emph{Casoli} has the potential to be promoted through a community-involved tourism approach.

We include four stakeholder communities in the evaluation of the response to the initiative. We do not speak about {\em success} since the initiative is deliberately soft, an outcome expected in years, and the definition of the evaluation criteria raises new questions. We are content with determining if the stakeholders viewed the initiative as intrusive, helpful, or simply neutral.

@@

\begin{itemize}
	\item {\em the residents} the local community that currently inhabits the site
	\item {\em the entrepreneurs} who currently have a business on the site
	\item {\em the administrators} that are in charge of managing the site resources and that, at due time, will respond to the two stakeholders above, and finally
	\item {\em the users}, those that come to \emph{Casoli} to visit the Cave of \emph{La Piella} and that we find the QR-codes on their way
\end{itemize}
